Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2013-11-11 Dail_Mad_ext.tex
intro edits

Theoretical ecology requires models of population dynamics for testing
hypotheses regarding spatial and temporal variation in
abundance. For example, much theoretical work has focused on
understanding the importance and existence of phenomenon such as
density-dependent population regulation, population cycling, and spatial synchrony
\citep{may:1975,turchin:1990,bjornstad_etal:1999}, %,royama:1977,dennis_taper:1994
and population models are required to evaluate associated hypotheses.
In applied contexts, population models are used for estimating extinction probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011,hostetler_etal:2012} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.
In order to address these questions, two complicating factors must
be confronted when fitting population models to data. 
First, deterministic models of population dynamics are virtually always inadequate due to process
variation, the inherent stochasticity in demographic parameters and environmental
conditions \citep{bjornstad_grenfell:2001,saether_engen:2002}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{link_nichols:1994,kery_etal:2009}.

State-space models are a widely used approach for
studying population dynamics while accounting for process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_etal:2006}. Classical state-space
models are time-series models in which the true state of the
system (e.g. population size during each year) is observed
imperfectly. One reason for the widespread adoption of
state-space models in ecology is that failure to account for process
variation and observation error can bias estimators of abundance and
population growth parameters. For instance, the strength of
density dependence will be over-estimated if observation error is
ignored \citep{link_nichols:1994,shenk_etal:1998}.

A simple state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\eta_t$.
% Richard, is switching this from $\grave{o}$ to $\eta$ okay?  Andy thought
% the o with accent was weird and I think I agree.  $\nu$ seems standard,
% but we use it for a slightly different meaning (ES only) later.  
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model, or it might be density-independent, such as when growth
is exponential.
%, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$ is the intrinsic rate of increase.
The full model can now be written as:
\begin{subequations}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \label{eq:ss1a} \\
N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \;
t=2,\hdots,T \label{eq:ss1b} \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T \label{eq:ss1c}
  \end{align}
\end{subequations}
where $\eta_t$ is the random effect allowing for process
variation unaccounted for by the deterministic model. In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian deviates:
$\eta_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. Standard practice is to
ignore process variation associated with $N_1$, as
indicated by Eq~\ref{eq:ss1a}. Alternatively, the population is
assumed to be at equilibrium such that the $N_1$ can be regarded as an
outcome of the equilibrium distribution.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
are among the most widespread approaches for modeling population dynamics
using time-series data, several problems exist. First,
time-series data and the underlying abundance parameters of interest
are typically integer-valued, as in the case of count data, raising
concerns about the use of Gaussian distribution for the random
effects. Second, the model for observation error has little biological basis. 
Third, the model does not account for spatial variation in abundance. Lastly, some of the
parameters of the model are not estimable. We briefly discuss each of
these points before describing a general approach to resolving these
shortcomings.

Use of the Gaussian distribution for modeling random process variation
and observation error is motivated by convenience rather than
biology. Specifically, the Gaussian assumptions
allow for parameter estimation using the Kalman filter
\citep{dennis_etal:2006}, which is much more computationally efficient
than estimation methods when random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}. %The problem with this is that it
This approach allows for negative and non-integer values of the state-variable, which is
inconsistent with the observed data, and may result in implausible predictions.
Transforming the count data and the underlying abundance variables do
little to resolve these issues \citep{ohara_kotze:2010}.

Another problem with standard state-space models is that the
observation error has no clear interpretation. The use of a mean zero
Gaussian distribution implies that $X_t$ will be
higher than $N_t$ as often as it is lower than $N_t$. 
Identifying a biological mechanism that would cause such symmetric errors
is difficult.
A more likely form of observation error, and one that has been recognized for well
over a century, results from failing to detect individuals that are
present. Imperfect detection may be attributable to
characteristics of the species under study, such as its elusiveness,
or to individuals being missed by the ecologist collecting the data in the field.
Although a vast number of methods have been devised for accounting for
this form of observation error, rarely have these methods been
integrated into state-space models \citep[but
see][]{buckland_etal:2004}.

@@

A more serious problem %than the ones associated with the Gaussian assumptions 
is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$, % or $X_1$;
although $N_1$ is a random variable. Thus, uncertainty exists
that should be accounted for in the model.  
%are needed to fully specify the model.
To be more specific, a fully-specified state-space model requires at least
three probability distributions, which we represent
using bracket notation (i.e., $[Y|\Omega]$ is interpreted
as the probability distribution of the random variable $Y$ given the
parameter $\Omega$). The three probability distributions required for a
state-space model are:
%%%% JEFF, does this extra sentence help? This notation comes from
%%%% several of the original state-space papers and is widespread in
%%%% books like Andy's and Bill's.
%\hl{[I'm not familiar with bracket notation,
% so these equations are confusing to me. How widely used is this
% notation in ecology? Will others likely understand this? Also, we might want
%to save space by trimming a bunch of this discussion back down.]}
\begin{subequations}
  \label{eq:ss2}
  \begin{align}
    [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
[N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\
\label{eq:ss2c}
    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
  \end{align}
\end{subequations}
where $\bm{\theta}$ are the process variation parameters for the
initial state, $\bm{\Theta}$ are the process variation parameters
influencing how abundance changes over time, and $\bm{p}$ are the
observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
models, but what should be the distribution for Eq.~\ref{eq:ss2a}?
%The three lines of general model
%described by Eq.~\ref{eq:ss2} correspond to the three equations shown
%for the specific example in Eq~\ref{eq:ss1}. However, in this case,
%we allow $N_1$ to be a random variable. But what should its
%distribution, $[N_1|\bm{\theta}]$ be?
And how can $\bm{\theta}$ be estimated given that there is only a single observation
available? 
%Often, these problems are simply assumed away. For
%example, sometimes researchers assume that there is no process variation in
%the first year, although this seems unjustified. Alternatively, the
%population may be assumed to be at equilibrium such that the expected
%value of $N_t$ is constant through time. Although this 
%makes the parameters $\bm{\theta}$ identifiable, assuming equilibrium
%defeats the objective of many studies of population dynamics, namely
%determining why a population varies over time.
%% OR
%Sometimes researchers assume that there is no process variation in 
%the first year or that the population is at equilibrium such that the expected
%value of $N_t$ is constant through time, but these approaches are unsatisfactory.

Several extensions of state-space models have been proposed to
overcome the limitations described above. \citet{devalpine_hastings:2002} and
\citet{kery_etal:2009} described methods for fitting models with non-Gaussian
distributions for the process and observation errors. Observation models with
more intuitive interpretations, such as those that explicitly model
detection probability, have been proposed by \citet{kery_etal:2009}. \citet{lele_etal:1998} and
\citet{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, only
the work by \citet{kery_etal:2009} addressed several of these limitations
simultaneously. However, their model did not include serial
dependence, which is a hallmark of population models. This
limits the utility of their model for making inferences about explicit
population processes.

In this paper, we focus on the model of \citet[henceforth the DM model]{dail_madsen:2011}
that simultaneous resolves each of the aforementioned problems with
traditional state-space models, and is designed for simple count data.
In the following section, we describe the DM model in its original
form and explain how it resolves each of the deficiencies with
standard state-space models. In Section~\ref{sec:ext}, we extend the
model to accommodate classical models of population growth and
to handle several features common to ecological
time-series. Specifically, we describe methods for accommodating
excess zeros and nuisance variables such as random observer effects.
Both frequentist and Bayesian methods of inference are discussed, and
code for fitting models is presented in the appendices.
%The Bayesian method
%is attractive in that it can accommodate prior information about
%detection parameters, which is available for many species, yet not
%collected as part of many existing monitoring programs.
In Section~\ref{sec:app}, we evaluate the performance of the model
extensions using a simulation study and by analyzing data from the
North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}. The overarching aim
of the paper is to provide ecologists with flexible and accessible means of
addressing important questions related to the variation of abundance in
space and time. 
%believe these extensions will increase the utility of the model for
%application to existing datasets such the Breeding Bird Survey.
\begin{comment}
  The DM model allows for inference about spatial and temporal
dynamics in abundance while accounting for observation error
using only spatially and temporally replicated count data. Their model
is an extension of the closed-population $N$-mixture model
\citep{royle:2004biom}, which was designed specifically to
address the problem of modeling spatial variation in abundance in the
  presence of observation error, specifically, when detection
  probability is less than one. The DM extension relaxes the
assumption of population closure and includes explicit
parameters describing population change as a first-order Markovian
  process. [more details]

>>>>

%Theoretical ecology requires models of population dynamics for testing
%hypotheses regarding spatial and temporal variation in
%abundance. For example, much theoretical work has focused on
Models of population dynamics are vital in both theoretical and
applied ecological research. For instance, %the have been used to assess
the importance and existence of phenomenon such as
density-dependent population regulation, population cycling, and
spatial synchrony have been studied by comparing data from natural
populations with theoretical models
\citep{may:1975,turchin:1990,bjornstad_etal:1999,royama:1977}. %,dennis_taper:1994
%and population models are required to evaluate associated hypotheses.
In applied contexts, population models are used for estimating extinction probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011,hostetler_etal:2012} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.

Several challenges are routinely encountered when trying to fit models
of population dynamics to data from field studies. % is that 
%In order to address these questions, two complicating factors must
%be confronted when fitting population models to data. 
First, deterministic models of population dynamics are virtually always inadequate due to process
variation, the inherent stochasticity in demographic parameters and environmental
conditions \citep{bjornstad_grenfell:2001,saether_engen:2002}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{link_nichols:1994,kery_etal:2009}.

Recently developed state-space models %are a widely used approach for
have made it possible to %become the primary tool for 
study population dynamics while accounting for both process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_etal:2006}. Classical state-space
models are time-series models in which the true state of the
system (e.g. population size during each year) is modeled as a hidden
process, and the observed data are modeled conditional on this process
and the observation error. One reason for the widespread adoption of
state-space models in ecology is that failure to account for 
%process variation and 
observation error can bias estimators of abundance and
population growth parameters. For instance, the strength of
density dependence will be over-estimated if observation error is
ignored \citep{link_nichols:1994,shenk_etal:1998}.

A simple state-space model can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\eta_t$.
% Richard, is switching this from $\grave{o}$ to $\eta$ okay?  Andy thought
% the o with accent was weird and I think I agree.  $\nu$ seems standard,
% but we use it for a slightly different meaning (ES only) later.  
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model, or it might be density-independent, such as when growth
is exponential.
%, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$ is the intrinsic rate of increase.
A generic model can be written: %The full model can now be written as:
\begin{subequations}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \label{eq:ss1a} \\
N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \;
t=2,\hdots,T \label{eq:ss1b} \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T. \label{eq:ss1c}
  \end{align}
\end{subequations}
%where $\eta_t$ is the random effect allowing for process
%variation unaccounted for by the deterministic model. 
In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian deviates:
$\eta_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. Standard practice is to
ignore process variation associated with $N_1$, as
indicated by Eq~\ref{eq:ss1a}. Alternatively, the population is
assumed to be at equilibrium such that the $N_1$ can be regarded as an
outcome of the equilibrium distribution.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
are among the most widespread approaches for modeling population dynamics
using time-series data, several problems exist. First,
time-series data and the underlying abundance parameters of interest
are typically integer-valued, as in the case of count data, raising
concerns about the Gaussian assumption for the random
effects. Second, the model for observation error is implausible. %has little biological basis. 
Third, the model does not account for spatial variation in abundance. Lastly, some of the
parameters of the model are not estimable. We briefly discuss each of
these points before describing a general approach to resolving these
shortcomings.

%Use of the Gaussian distribution for modeling random process variation
%and observation error is often motivated by convenience rather than
%biology. Specifically, t
The assumption that random effects are Gaussian distributed makes it
easier to estimate parameters because methods such as the Kalman
filter can be applied %The Gaussian assumption allows for efficient methods of parameter
%estimation, such as the Kalman filter
\citep{dennis_etal:2006}, which are not available otherwise. % which is much more computationally efficient
%than estimation methods when random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}. %The problem with this is that it
However, these models effectively assume that abundance can be take on any
real value, including negative values, 
%This approach allows for negative and non-integer values of the state-variable, which is
%inconsistent with the observed data, and may 
and these models can result in implausible predictions.
Transforming the count data and the underlying abundance variables 
do little to resolve these issues \citep{ohara_kotze:2010}.

Another problem with standard state-space models is that the model for
the mean zero Gaussian distribution for the observation error
implies that false positive detections are
equally likely as false negatives. To our knowledge, all studies to
date have shown that false negative detections are much more common
that false positives (CITE DAVE MILLER PAPERS). 
%will be
%higher than $N_t$ as often as it is lower than $N_t$. 
%Identifying a biological mechanism that would cause such symmetric errors
%is difficult.
A more likely form of observation error, and one that has been recognized for well
over a century, results from failing to detect individuals that are
present. Imperfect detection may be attributable to
characteristics of the species under study, such as its elusiveness,
or to individuals being missed by the ecologist collecting the data in the field.
Although a vast number of methods have been devised for accounting for
this form of observation error, rarely have these methods been
integrated into state-space models \citep[but
see][]{buckland_etal:2004}.

@@

A more serious problem %than the ones associated with the Gaussian assumptions 
is that the parameters of simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Specifically, abundance in
the first year cannot be estimated because there is only one data
point (the count in the first year) that carries any information about
it. 
%Recall that the
%Eq.~\ref{eq:ss1} did not specify distributions for $N_1$, % or $X_1$;
%although $N_1$ is a random variable. Thus, uncertainty exists
%that should be accounted for in the model.  
%%are needed to fully specify the model.
%To be more specific, a fully-specified state-space model requires at least
%three probability distributions, which we represent
%using bracket notation (i.e., $[Y|\Omega]$ is interpreted
%as the probability distribution of the random variable $Y$ given the
%parameter $\Omega$). The three probability distributions required for a
%state-space model are:
%%%% JEFF, does this extra sentence help? This notation comes from
%%%% several of the original state-space papers and is widespread in
%%%% books like Andy's and Bill's.
%\hl{[I'm not familiar with bracket notation,
% so these equations are confusing to me. How widely used is this
% notation in ecology? Will others likely understand this? Also, we might want
%to save space by trimming a bunch of this discussion back down.]}
%\begin{subequations}
%  \label{eq:ss2}
%  \begin{align}
%    [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
%[N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\
%\label{eq:ss2c}
%    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
%  \end{align}
%\end{subequations}
%where $\bm{\theta}$ are the process variation parameters for the
%initial state, $\bm{\Theta}$ are the process variation parameters
%influencing how abundance changes over time, and $\bm{p}$ are the
%observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
%and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
%models, but what should be the distribution for Eq.~\ref{eq:ss2a}?
%The three lines of general model
%described by Eq.~\ref{eq:ss2} correspond to the three equations shown
%for the specific example in Eq~\ref{eq:ss1}. However, in this case,
%we allow $N_1$ to be a random variable. But what should its
%distribution, $[N_1|\bm{\theta}]$ be?
%And how can $\bm{\theta}$ be estimated given that there is only a single observation
%vailable? 
%Often, these problems are simply assumed away. For
%example, sometimes researchers assume that there is no process variation in
%the first year, although this seems unjustified. Alternatively, the
%population may be assumed to be at equilibrium such that the expected
%value of $N_t$ is constant through time. Although this 
%makes the parameters $\bm{\theta}$ identifiable, assuming equilibrium
%defeats the objective of many studies of population dynamics, namely
%determining why a population varies over time.
%% OR
%Sometimes researchers assume that there is no process variation in 
%the first year or that the population is at equilibrium such that the expected
%value of $N_t$ is constant through time, but these approaches are unsatisfactory.

Several extensions of state-space models have been proposed to
overcome the limitations described above. \citet{devalpine_hastings:2002} and
\citet{brooks_etal:2004} described methods for fitting models with non-Gaussian
distributions for the process and observation errors. Observation models with
more intuitive interpretations, such as those that explicitly model
detection probability, have been proposed by 
\citet{kery_etal:2009}. \citet{lele_etal:1998} and 
\citet{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, only
the work by \citet{kery_etal:2009} addressed several of these limitations
simultaneously; however, their model did not include serial
dependence, which is a hallmark of population models. %This
%limits the utility of their model for making inferences about explicit
%population processes.

In this paper, we focus on the model of \citet[henceforth the DM model]{dail_madsen:2011}
that resolves many of the problems with classical state-space models.
%simultaneous resolves each of the aforementioned problems with
%traditional state-space models, and is designed for simple count data.
%In the following section, we describe the DM model in its original
%form and explain how it resolves each of the deficiencies with
%standard state-space models. 
In Section~\ref{sec:ext}, we extend the
model to accommodate classical models of population growth and
to handle several features common to ecological
time-series, such as excess zeros and unexplained random variation. 
%Specifically, we describe methods for accommodating
%excess zeros and nuisance variables such as random observer effects.
Both frequentist and Bayesian methods of inference are discussed, and
code for fitting models is presented in the appendices.
%The Bayesian method
%is attractive in that it can accommodate prior information about
%detection parameters, which is available for many species, yet not
%collected as part of many existing monitoring programs.
In Section~\ref{sec:app}, we evaluate the performance of the model
extensions using a simulation study and by analyzing data from the
North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}. The overarching aim
of the paper is to provide ecologists with flexible and accessible means of
addressing important questions related to the variation of abundance in
space and time. 
%believe these extensions will increase the utility of the model for
%application to existing datasets such the Breeding Bird Survey.
\begin{comment}
  The DM model allows for inference about spatial and temporal
dynamics in abundance while accounting for observation error
using only spatially and temporally replicated count data. Their model
is an extension of the closed-population $N$-mixture model
\citep{royle:2004biom}, which was designed specifically to
address the problem of modeling spatial variation in abundance in the
  presence of observation error, specifically, when detection
  probability is less than one. The DM extension relaxes the
assumption of population closure and includes explicit
parameters describing population change as a first-order Markovian
  process. [more details]
